Evolution Reinforces Cooperation with the Emergence of Self-Recognition Mechanisms: an Empirical Study of the Moran process for the Iterated Prisoner's Dilemma

### 2017-10-30 main.tex
Include better explanation of the Moran details

Recent work has argued that agent-based
simulations can provide insights in evolutionary game theory not available
via direct mathematical analysis \cite{adami2016evolutionary}. The results
and insights contained in this paper would be difficult to derive analytically.

@@

Figure~\ref{fig:moran_process} shows a diagrammatic representation of the Moran
process, a stochastic birth death process on a finite population in which the
population size stays constant over time. Individuals are \textbf{selected}
according to a given fitness landscape. Once selected, the individual is
reproduced and similarly another individual is chosen to be removed from the
population. In some settings mutation is also considered but without mutation
(the case considered in this work) this process will arrive at an absorbing
state where the population is entirely made up of players of one strategy. The
probability with which a given strategy takes over a population is called the
\textit{fixation probability}. A more detailed analytic description of this is
given in Section~\ref{sec:validation}. In our simulations offspring do not
inherit any knowledge or history from parent replicants.

@@

Apart from the PSO Gambler strategy, which was trained using a particle swarm
optimisation algorithm, these strategies are trained with an evolutionary
algorithm that perturbs strategy parameters and optimizes the mean total score
against all other opponents \cite{affenzeller2009genetic}. They were trained to
win IPD tournaments by maximizing their mean total payoffs against a variety
of opponents. Variation is
introduced via mutation and crossover of parameters, and the best performing
strategies are carried to the next generation along with new variants. Similar
methods appear in the literature \cite{Ashlock2006}.

@@

There are three further strategies trained specifically for this study; Trained
FSM 1, 2, and 3 (TF1 - TF3). These are based on finite state machines of 16, 16,
and 8 states respectively (see Figures~\ref{fig:tf1},~\ref{fig:tf2}
and~\ref{fig:tf3}).
% TODO Describe what the diagram shows

\begin{figure}[!hbtp]
    \centering
    \scalebox{.5}{\input{tex/fsm_one.tex}}
    \caption{TF1: a 16 state finite state machine with a handshake leading to
    mutual cooperation at state 4.}
    \label{fig:tf1}
\end{figure}

\begin{figure}[!hbtp]
    \centering
    \scalebox{.5}{\input{tex/fsm_two.tex}}
    \caption{TF2: a 16 state finite state machine with a handshake leading to
    mutual cooperation at state 16.}
    \label{fig:tf2}
\end{figure}

@@

Each strategy pair is run for 1000 repetitions of the Moran process to fixation
with starting population distributions of $(1, N-1)$, $(N/2, N/2)$ and $(N-1 ,
1)$, for \(N\) from 2 through 14. The fixation probability is then empirically
computed for each combination of starting distribution and value of \(N\).  The
Axelrod library can carry out exact simulations of the Moran process. Since some
of the strategies have a high computational cost or are stochastic, samples are
% TODO Add a bit more clarity that the cache is indeed stochastic (although not
% when using deterministic strategies.
taken from a large number of match outcomes for the pairs of players for use in
computing fitnesses in the Moran process. This approach was verified to agree
with unsampled calculations to a high degree of accuracy in specific cases.
This is described in Algorithms~\ref{alg:data_collection}
and~\ref{alg:moran_process}.
% TODO Are we clear about selection procedure?
% TODO Specify that kill off is uniformly random.

@@

\begin{algorithm}[!hbtp]
        \caption{Moran process}
        \label{alg:moran_process}
        \input{tex/moran_process_simulation.tex}
        % TODO Add description of number of turns in the IPD
\end{algorithm}

>>>>

Recent work has argued that agent-based
simulations can provide insights in evolutionary game theory not available
via direct mathematical analysis~\cite{adami2016evolutionary}. The results
and insights contained in this paper would be difficult to derive analytically.

@@

Figure~\ref{fig:moran_process} shows a diagrammatic representation of the Moran
process, a stochastic birth death process on a finite population in which the
population size stays constant over time. Individuals are \textbf{selected}
according to a given fitness landscape. Once selected, the individual is
reproduced and similarly another individual is chosen to be removed from the
population (a uniform random selection is used). In some settings mutation is 
also considered but without mutation
(the case considered in this work) this process will arrive at an absorbing
state where the population is entirely made up of players of one strategy. The
probability with which a given strategy takes over a population is called the
\textit{fixation probability}. A more detailed analytic description of this is
given in Section~\ref{sec:validation}. In our simulations offspring do not
inherit any knowledge or history from parent replicants.

@@

Apart from the PSO Gambler strategy, which was trained using a particle swarm
optimisation algorithm, these strategies are trained with an evolutionary
algorithm that perturbs strategy parameters and optimizes the mean total score
against all other opponents \cite{affenzeller2009genetic}. They were trained to
win IPD tournaments by maximizing their mean total payoffs against a variety
of opponents. Variation is
introduced via mutation and crossover of parameters, and the best performing
strategies are carried to the next generation along with new variants. Similar
methods appear in the literature \cite{Ashlock2006}.
There has also been some work on strategies using an evolutionary in real time: 
in~\cite{Gaudesi2016}
an evolutionary algorithm is used to build a model of the opponent and attempt
to exploit any potential weakness. This sort of work is not considered here:
strategies are instead pre trained.

@@

There are three further strategies trained specifically for this study; Trained
FSM 1, 2, and 3 (TF1 - TF3). These are based on finite state machines of 16, 16,
and 8 states respectively. These are shown in Figures~\ref{fig:tf1},~\ref{fig:tf2}
and~\ref{fig:tf3}), using the notation common in the literature where
\(A_1/A_2\) is the action of the opponent \(A_1\) and the response of the player
\(A_2\) as well as arrows corresponding to changes of state.

\begin{figure}[!hbtp]
    \centering
    \scalebox{.7}{\input{tex/fsm_one.tex}}
    \caption{TF1: a 16 state finite state machine with a handshake leading to
    mutual cooperation at state 4.}
    \label{fig:tf1}
\end{figure}

\begin{figure}[!hbtp]
    \centering
    \scalebox{.7}{\input{tex/fsm_two.tex}}
    \caption{TF2: a 16 state finite state machine with a handshake leading to
    mutual cooperation at state 16.}
    \label{fig:tf2}
\end{figure}

@@

Each strategy pair is run for 1000 repetitions of the Moran process to fixation
with starting population distributions of $(1, N-1)$, $(N/2, N/2)$ and $(N-1 ,
1)$, for \(N\) from 2 through 14. The fixation probability is then empirically
computed for each combination of starting distribution and value of \(N\).  The
Axelrod library can carry out exact simulations of the Moran process. Since some
of the strategies have a high computational cost or are stochastic, samples are
taken from a large number of 200 turn match outcomes for the pairs of players
for use in computing fitnesses in the Moran process (i.e. a stochastic cache of
matches is used). This approach was verified to agree with unsampled
calculations to a high degree of accuracy in specific cases.  This is described
in Algorithms~\ref{alg:data_collection} and~\ref{alg:moran_process}.

@@

\begin{algorithm}[!hbtp]
        \caption{Moran process}
        \label{alg:moran_process}
        \input{tex/moran_process_simulation.tex}
\end{algorithm}
